Equalizer Programming Guide

### 2009-07-27 Developer/ProgrammingGuide/paper.tex
Programming Guide review pass
[ ] Will break Win32 build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

\date{
  \vspace{15cm}
  {\Large\sffamily The official reference for creating parallel,
    scalable OpenGL\texttrademark\ applications with the Equalizer
    parallel rendering framework}\\\vspace{1cm}
  Version 1.5.8 for Equalizer 0.6+svn\\[\medskipamount]
%  \today
}

@@

\clearpage
\pagenumbering{roman}
\pagestyle{headings}
\tableofcontents
\listoffigures \vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Revision for Equalizer 0.5\\
    1.4     & Nov 25, 2008 & Revision for Equalizer 0.6\\
    1.5.1   & Dec 08, 2008 & Minor updates based on the print version\\
    1.5.2   & Jan 10, 2009 & Added \sref{sCompositor}\\
    1.5.3   & Jan 13, 2009 & Added \sref{sNetwork}\\
    1.5.4   & Feb 09, 2009 & Added \sref{sNetUsage}\\
    1.5.5   & Mar 10, 2009 & Updated to reflect new AGL event dispatch\\
    1.5.6   & Mar 23, 2009 & Documented Layout API\\
    1.5.7   & Apr 17, 2009 & Updated Layout API (multiple observers)\\
    1.5.8   & May 30, 2009 & Added \sref{sEqualizers}, updated \sref{sLoadBalancing} and \sref{sSwapBarrier}\\
    \hline
  \end{tabularx}}
\clearpage

@@

Some scalable rendering parameters have to be updated at run-time to
provide optimal performance or influence other scalable rendering
features. These adjustments are often referred to as load-balancing, but
are called \textsf{equalizers}, since their functionality is not limited
to load-balancing in Equalizer.

The server provides a number of equalizers, which automatically update
certain parameters of compounds based on run-time information. They
balance the load of 2D and DB compounds, optimally distribute render
resources for multi-view configurations, adapt the resolution to provide
a constant framerate or zoom images to allow monitoring of another view.

@@

Equalizer uses a C++ programming interface. The API is minimally
invasive, so Equalizer imposes only a minimal, natural execution
framework upon the application. It does not provide a scene graph, or
interfere in any other way with the application's rendering code. The
restructuring work required for Equalizer is the minimal refactoring
needed to parallelize the application for rendering.

All rendering threads in the configuration run asynchronously to the
application's main thread. Depending on the configuration's latency,
they can fall $n$ frames behind the last frame finished by the
application thread. A latency of one frame is usually not perceived by
the user, but can increase rendering performance substantially.

\subsection{\label{sConfig}Config}

The \textsf{eq::Config} represents the current configuration of the
application. The configuration is the session in which all render
clients are registered. A configuration consists of the description of
the rendering resources and the usage description for these resources.

The rendering resources are represented in a hierarchical tree structure
which corresponds to the physical and logical resources found in a 3D
rendering environment.

@@

@@

\subsubsection{Node}

@@

\subsubsection{Pipe}

@@

\subsubsection{Window}

The \textsf{eq::Window} class encapsulates a drawable and an OpenGL
context. The drawable can be an on-screen window or an off-screen
PBuffer. FrameBuffer object (FBO) support will be implement by a later
version of Equalizer. The window uses an \textsf{eq::OSWindow}, which
abstracts and manages window-system-specific handles to the drawable and
context, e.g., an X11 window \textsf{XID} and \textsf{GLXContext} for
the glX window system.

\subsubsection{Channel}

@@

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/canvas.pdf}
  {\caption{\label{fCanvas}A canvas using four, non-consecutive segments}}
\end{wrapfigure}
A segment represents one output channel of the canvas, e.g., a projector
or a LCD. A segment has an output channel, which references the channel to
which the display device is connected.

@@

Passive stereo installations have one segment per eye pass, where the
segment for the left and right eye fully overlap.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/layout.png}
  {\caption{\label{fLayout}A layout with four views}}
  \vspace{-4ex}
\end{wrapfigure}
A layout is the grouping of logical views. It is used by one or more
canvases. For all given layouts, Equalizer creates destination channels
when the configuration file is loaded. These destination channels can be
referenced by compounds to configure scalable rendering.

@@

@@

@@

@@

\subsubsection{\label{sFrustum}Frustum}
\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/frusta.pdf}
  {\caption{\label{fFrusta}Wall and projection parameters}}
\end{wrapfigure}
Compounds have a frustum description to define the physical layout of
the display environment.

@@

The frustum can be specified as a wall or projection description.

The projection is defined by the position and head-pitch-roll
orientation of the projector, as well as the horizontal and vertical
field-of-view and distance of the projection wall.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/viewLB.pdf}
  {\caption{\label{fViewLoadBalancing}\small Cross-segment
      load-balancing for two segments using eight GPU's}}
\end{wrapfigure}
\fig{fViewLoadBalancing}\footnote{Image Copyright Realtime Technology
  AG, 2008} illustrates this process. On the left side, a static
assigment of resources to display segments is used. The right-hand
segment has a higher load than the left-hand segment, causing
sub-optimal performance. The configuration on the left uses a view
equalizer, which assigns two GPU's to the left segment and four GPU's to
the right segment, which leads to optimal performance for this model and
camera position.

The view equalizer can also use resources from another display resource,
if this resource has little rendering load by itself. It is therefore
possible to improve the rendering performance of a multi-display system
without any additional resources. This is particulary useful for
installations with a higher number of displays where the rendering load
is typically in a few segments only, e.g., for a CAVE.

@@

\paragraph{DFR Equalizer}
Dynamic Frame Resolution (DFR) trades rendering performance for visual
quality. The rendering for a channel is done at a different resolution
than the native channel resolution to keep the framerate constant. The
DFR equalizer adapts the zoom of a channel, based on the target and
current framerate. It is typically used for fill-rate bound
applications, such as volume rendering and raytracing.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/dfr.pdf}
  {\caption{\label{fDFR}\small Dynamic Frame Resolution}}
\end{wrapfigure}
\fig{fDFR}\footnote{Data set courtesy of Olaf Ronneberger, Computer
  Science Institute, University of Freiburg, Germany} shows DFR for
volume rendering. To achieve 10 frames per second, the model is rendered
at a lower resolution, and upscaled to the native resolution for
display. The rendering quality is slightly degraded, while the rendering
performance remains interactive. When the application is idle, it
renders a full-resolution view.

@@

@@

In this section the source code of \textsf{eqPly} is explained in
detail, and relevant design decisions, caveats and other remarks are
raised.

@@

The distributed data classes are helper classes based on
\textsf{eq::net::Object}. They illustrate the typical usage of distributed
objects for static as well as dynamic, frame-specific data. Furthermore
they are used for a basic scene graph distribution of the model data.

@@

The second step is to initialize the Equalizer library. The
initialization function of Equalizer also parses the command line, which
is used to set certain default values based on Equalizer-specific
options\footnote{Equalizer-specific options always start with -\,-eq-},
e.g., the default server address. Furthermore, a \textsf{NodeFactory} is
provided. The \textsf{EQERROR} macro, and its counterparts
\textsf{EQWARN}, \textsf{EQINFO} and \textsf{EQVERB} allow selective
debugging outputs with various logging levels:

@@

The third step is to create an instance of the application and to
initialize it locally. The application is a subclass of
\textsf{eq::Client}, which in turn is an \textsf{eq::net::Node}. The
underlying network layer in Equalizer is a peer-to-peer network of
\textsf{eq::net::Node}s. The client/server concept is implement the
higher-level \textsf{eq} client namespace.

The application programmer does not usually have to be aware of the
classes in the \textsf{eq::net} namespace, but both the
\textsf{eq::Client} and the Equalizer server are
\textsf{eq::net::Node}s.

@@

{\footnotesize\begin{lstlisting}
    // 3. initialization of local client node
    RefPtr< eqPly::Application > client = new eqPly::Application( initData );
    if( !client->initLocal( argc, argv ))
    {
        EQERROR << "Can't init client" << endl;
        eq::exit();
        return EXIT_FAILURE;
    }
\end{lstlisting}}%>>

@@

{\footnotesize\begin{lstlisting}
    // 4. run client
    const int ret = client->run();
\end{lstlisting}}

@@

{\footnotesize\begin{lstlisting}
    // 5. cleanup and exit
    client->exitLocal();
    client = 0;

    eq::exit();
    return ret;
}
\end{lstlisting}}

@@

\begin{enumerate}
\item \textbf{Application}: The executable started by the user, the
  controlling entity of the rendering session.
\item \textbf{Auto-launched render client}: The typical render client,
  started by the server. The server starts the executable with special
  parameters, which cause \textsf{Client::initLocal} to never
  return. During exit, the server terminates the process. By default,
  the server starts the render client using \textsf{ssh}. The launch
  command can be used to configure another program to auto-launch render
  clients.
\item \textbf{Resident render client}: Manually pre-started render
  client, listening on a specified port for server commands. This mode
  is selected using the command-line option \textsf{--eq-client} and
  potentially \textsf{--eq-listen \textless address\textgreater} and
  \textsf{-r}\footnote{see
    \link{http://www.equalizergraphics.com/documents/design/residentNodes.html}}.
\end{enumerate}

@@

@@

When the frame is started, the server computes all rendering tasks and
sends them to the appropriate render client nodes. The render client
nodes dispatch the tasks to the correct node or pipe thread, where they
are executed in order of arrival.

\textsf{Config::finishFrame} synchronizes on the completion of the frame
\textsf{current - latency}. The latency is specified in the
configuration file, and allows several outstanding frames. This allows
overlapping execution in the node and pipe threads and minimizes idle
times.

By default, \textsf{Config::finish\-Fra\-me} also synchronizes the
completion of all local rendering tasks for the current frame. This
facilitates porting of existing rendering codes, since the database does
not have to be multi-buffered. Applications such as \textsf{eqPly}, which
do not need this per-node frame synchronization can disable it, as
explained in \sref{sThreadModel}.

\begin{figure}[ht!]\center
  \includegraphics[width=\textwidth]{images/syncAsync}
  {\caption{\label{fSyncAsync}Synchronous and asynchronous execution}}
\end{figure}
\fig{fSyncAsync} shows the execution of the rendering tasks of a 2-node
2D compound without latency and with a latency of one frame. The
asynchronous execution pipelines certain rendering operations and hides
imbalances in the load distribution, resulting in an improved
framerate. With \textsf{eqPly}, a speedup of 15\% has been observed on a
five-node rendering cluster when using a latency of one frame instead of
no
latency\footnote{\link{http://www.equalizergraphics.com/scalability.html}}. A
latency of one or two frames is normally not perceived by the user.

@@

If the application detects that it is currently idle, all pending frames
are flushed. Then it waits and handles one event at a time, until a
redraw is needed:

@@

        if( !config->needsRedraw( ))
            config->finishAllFrames(); // flush, TODO flush one task at a time

        while( !config->needsRedraw( )) // wait for an event requiring redraw
        {
            const eq::ConfigEvent* event = config->nextEvent();
            if( !config->handleEvent( event ))
                EQVERB << "Unhandled " << event << endl;
        }
    }
\end{lstlisting}}%>>

@@

Equalizer provides distributed objects which help implementing data
distribution in a cluster environment. The master version of a
distributed object is registered with a \textsf{eq::net::Session}, which
assigns a session-unique identifier to the object. This identifier can
be used by other nodes to map their instance of the object, thus
synchronizing the object's data with the remotely registered master
version.

Distributed objects are created by subclassing from \textsf{eq::Object}
or \textsf{eq::net::Object}. Distributed objects can be static
(immutable) or dynamic. Dynamic objects are versioned.

@@

Distributed objects are addressed using session-unique identifiers,
because pointers to other objects cannot be distributed directly, since
they have no meaning on remote nodes. The session with which distributed
objects are registered is normally the \textsf{eq::Config}, which is
subclassed from \textsf{eq::net::Session}.

The shared data entry point for the render clients is the identifier
passed by the application to \textsf{Config::init}. This identifier
typically contains the identifier of a static distributed object, and is
passed by Equalizer to all \textsf{configInit} task methods. Normally
this initial object is mapped by the render clients, and it contains
identifiers to other shared data objects.

The distributed data objects referenced by the initial data object are
often versioned objects, to keep them in sync with the rendered
frames. Similar to the initial identifier passed to
\textsf{Config::init}, an object identifier or version can be passed to
\textsf{Config::startFrame}. Equalizer will pass this identifier to all
\textsf{frameStart} task methods. In \textsf{eqPly}, the frame-specific
data, e.g., the global camera data, is versioned. The frame data
identifier is passed in the initial data, and the frame data version is
passed with each new frame request.

@@

Equalizer determines the way changes are to be handled by calling
\textsf{Object::get\-Change\-Type} during the registration of the master
version of a distributed object. The change type determines the memory
footprint and the contract for calling the serialization methods. The
following change types are possible:

@@

Custom data type serializers can be implemented by providing the
appropriate serialization functions. No pointers should be directly
transmitted through the data streams. Typically the corresponding object
is a distributed object as well, and it's identifier and potentially
version is transmitted in place of a pointer.

For \textsf{InitData}, serialization in \textsf{getInstanceData}
and de-serialization in \textsf{applyInstanceData} is performed by
streaming all member variables to or from the provided data
streams. Efficient buffering and data transport between nodes is
implemented in the data streams:

{\footnotesize\begin{lstlisting}
void InitData::getInstanceData( eq::net::DataOStream& os )
{
    os << _frameDataID << _modelID << _windowSystem << _useVBOs << _useGLSL
       << _invFaces ;
}

void InitData::applyInstanceData( eq::net::DataIStream& is )
{
    is >> _frameDataID >> _modelID >> _windowSystem >> _useVBOs >> _useGLSL
       >> _invFaces;

@@

>>>>

\date{
  \vspace{15cm}
  {\Large\sffamily The official reference for creating parallel,
    scalable OpenGL\texttrademark\ applications with the Equalizer
    parallel rendering framework}\\\vspace{1cm}
  Version 1.5.9 for Equalizer 0.6+svn\\[\medskipamount]
%  \today
}

@@

\clearpage
\pagenumbering{roman}
\pagestyle{headings}
\tableofcontents
\listoffigures \vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Revision for Equalizer 0.5\\
    1.4     & Nov 25, 2008 & Revision for Equalizer 0.6\\
    1.5.1   & Dec 08, 2008 & Minor updates based on the print version\\
    1.5.2   & Jan 10, 2009 & Added \sref{sCompositor}\\
    1.5.3   & Jan 13, 2009 & Added \sref{sNetwork}\\
    1.5.4   & Feb 09, 2009 & Added \sref{sNetUsage}\\
    1.5.5   & Mar 10, 2009 & Updated to reflect new AGL event dispatch\\
    1.5.6   & Mar 23, 2009 & Documented Layout API\\
    1.5.7   & Apr 17, 2009 & Updated Layout API (multiple observers)\\
    1.5.8   & May 30, 2009 & Added \sref{sEqualizers}, updated \sref{sLoadBalancing} and \sref{sSwapBarrier}\\
    1.5.9.  & Jul 27, 2009 & Major review pass before release\\
    \hline
  \end{tabularx}}
\clearpage

@@

Some scalable rendering parameters are updated at run-time to provide
optimal performance or influence other scalable rendering
features. These adjustments are often referred to as load-balancing, but
are called \textsf{equalizers}, since their functionality is not limited
to load-balancing in Equalizer.

The server provides a number of equalizers, which automatically update
certain parameters of compounds based on run-time information. They
balance the load of 2D and DB compounds, optimally distribute render
resources for multi-segment projections systems, adapt the resolution to
provide a constant framerate or zoom images to allow monitoring of
another view.

@@

\section{\label{sConfig}Writing Configurations}

\subsection{Preparation}

Before writing a configuration, it is useful to assemble the following
information:

\begin{itemize}
\item A list of all \textbf{computers} in you rendering cluster,
  including the \textbf{IP addresses} of all network interfaces to be
  used.
\item The number of \textbf{graphics cards} in each computer.
\item The \textbf{physical dimensions} of the display system, if
  applicable. These are typically the bottom-left, bottom-right and
  top-left corner points of each display surfaces in meters.
\item The relative coordinates of all the \textbf{segments} belonging to
  each display surface, and the graphics card \textbf{output} used for
  each segment. For homogenous setups, it is often enough to know the
  number of rows and columns on each surface, as well as the overlap or
  underlap percentage, if applicable.
\item The number of desired \textbf{application windows}. Application
  windows are typically destination windows for scalable rendering or
  windows paired with a view on a display system.
\item \textbf{Characteristics} of the application, e.g., supported
  scalability modes and features.
\end{itemize}

\subsection{Overview}

Equalizer applications are configured at run-time by the Equalizer
server. The server loads its configuration from a text file, which is an
one-to-one representation of the configuration data structures at
runtime.

For an extensive documentation of the file format please refer to
\aref{aFileFormat}. This section gives an introduction on how to write
configuration files.

A configuration consists of the declaration of the rendering resources
and the usage description for these resources. It can optionally contain
the description of the physical layout of the projection system and
logical layouts on the projection canvases.

The rendering resources are represented in a hierarchical tree structure
which corresponds to the physical and logical resources found in a 3D
rendering environment: nodes (computers), pipes (graphics cards),
windows, channels.

@@

Physical layouts of display systems are configured using canvases with
segments, which represent 2D rendering areas composed of multiplex
displays or projectors. Logical layouts are applied to canvases and
define views on a canvas.

@@

For each machine in your cluster, create one \textsf{node}. Create one
\textsf{appNode} if your application process should also render, which
is often the case. List all nodes, even if you are not planning to use
them at first. Equalizer will only instantiate and access used nodes,
that is, nodes which are referenced by an active compound.

In each node, list all \textsf{connection}s thorugh which this node is
reachable. Typically a node uses only one connection, but it is possible
to configure multiple connections if the machine and cluster is set up
to use multiple, independent network interfaces. Make sure the
\textsf{hostname} used is reachable from all nodes. An IP address may be
used as the hostname.

@@

For each node, create on \textsf{pipe} for each graphics card in the
machine. Set the \textsf{device} number to the correct index. On
operating systems using X11, e.g., Linux, also set the port number if
your X-Server is running on a non-standard port.

@@

\subsection{Window}

Configure one \textsf{window} for each desired application window on the
appNode. Configure one full-screen window for each display
segment. Configure one off-screen window, typically a pbuffer, for each
graphics card used as a source for scalable rendering.

Sometimes display segments cover only a part of the graphics card
output. In this case it is adviced to configure a non-fullscreen window
without window decorations, using the correct window viewport.

The \textsf{eq::Window} class encapsulates a drawable and an OpenGL
context. The drawable can be an on-screen window or an off-screen
PBuffer or FrameBuffer Object (FBO). The window uses an
\textsf{eq::OSWindow}, which abstracts and manages
window-system-specific handles to the drawable and context, e.g., an X11
window \textsf{XID} and \textsf{GLXContext} for the glX window system.

Configure one \textsf{channel} for each desired rendering in each window
area. Typically one full-screen channel per window is used. Name the
channel using a unique, easily-identifiable name.

Multiple channels in application windows may be used to view the model
from different viewports.

Sometimes, a single window is split across multiple projectors, e.g., by
using an external splitter such as the Matrox triple-head-to-go. In this
case configure one channel for each segment, using the channel's
\textsf{viewport} to configure its position wit respect to the window.

@@

If you are writing a configuration for workstation usage you can skip
the following sections and restart with \sref{sCompounds}.

Configure one \textsf{canvas} for each display surface. For planar
surfaces, e.g., a display wall, configure a frustum. For non-planar
surfaces, the frustum will be configured on each display segment.

The frustum can be specified as a wall or projection description. Take
care to choose your reference system for describing the frustum to be
the same system as used by the head-tracking matrix calculated by the
application.

A projection is defined by the position and head-pitch-roll orientation
of the projector, as well as the horizontal and vertical field-of-view
and distance of the projection wall.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/frusta.pdf}
  {\caption{\label{fFrusta}Wall and projection parameters}}
\end{wrapfigure}

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/canvas.pdf}
  {\caption{\label{fCanvas}A canvas using four, non-consecutive segments}}
\end{wrapfigure}

Configure one \textsf{segment} for each display or projector of each
canvas. Configure the \textsf{viewport} of the segment to match the area
covered by the segment on the physical canvas. Set the output
\textsf{channel} to the resource driving the described projector.

For non-planar displays, configure the frustum as described in
\sref{sCanvas}. For passive stereo installations, configure one segment
per eye pass, where the segment for the left and right eye fully
have the same viewport.

A segment represents one output channel of the canvas, e.g., a projector
or a LCD. A segment has an output channel, which references the channel to
which the display device is connected.

@@

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/layout.png}
  {\caption{\label{fLayout}A layout with four views}}
\end{wrapfigure}

Configure one \textsf{layout} for each configuration of logical
views. Name the layout using a unique name. Often only one layout with a
one view is used for all canvases.

Enable the layout on each desired canvas by adding it to the
canvas. Since canvases reference layouts by name or index, layouts have
to be configured before their respective canvases in the configuration
file.

A layout is the grouping of logical views. It is used by one or more
canvases. For all given layout/canvas combinations, Equalizer creates
destination channels when the configuration file is loaded. These
destination channels can be referenced by compounds to configure
scalable rendering.

@@

Configure one \textsf{view} for each logical view in each layout. Set
the \textsf{viewport} to position the view.

@@

Unless you have multiple tracked persons, or want to disable tracking on
certain views, you can skip this section.

Configure one \textsf{observer} for each tracked person in the
configuration. Most configurations have at most one observer. Assign the
observer to all views which belong to this observer. Since the observer
is referenced by its name or index, it has to be specified before the
layout in the configuration file.

Views with no observer are not tracked. The config file loader will
create one default observer and assign it to all views if the
configuration has no observer.

@@

Compound trees are used to describe how the rendering resources are
combined to produce the desired output, and how they are aggregated to
increase the performance. It is adviced to study and understand the
basic configuration files shipped with the Equalizer configuration,
before attempting to write compound configurations.

For configurations using canvases and layouts without scalability, the
configuration file loader will create the appropriate compounds.

@@

\subsubsection{\label{sFrustum}Frustum}

Compounds have a frustum description to define the physical layout of
the display environment.

@@

The frustum can be specified as described in \sref{sCanvas}.

@@

\fig{fViewLoadBalancing}\footnote{Image Copyright Realtime Technology
  AG, 2008} illustrates this process. On the left side, a static
assigment of resources to display segments is used. The right-hand
segment has a higher load than the left-hand segment, causing
sub-optimal performance. The configuration on the left uses a view
equalizer, which assigns two GPU's to the left segment and four GPU's to
the right segment, which leads to optimal performance for this model and
camera position.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/viewLB.pdf}
  {\caption{\label{fViewLoadBalancing}\small Cross-segment
      load-balancing for two segments using eight GPU's}}
\end{wrapfigure}
The view equalizer can also use resources from another display resource,
if this resource has little rendering load by itself. It is therefore
possible to improve the rendering performance of a multi-display system
without any additional resources. This is particulary useful for
installations with a higher number of displays where the rendering load
is typically in a few segments only, e.g., for a CAVE.

@@

\paragraph{DFR Equalizer}
\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/dfr.pdf}
  {\caption{\label{fDFR}\small Dynamic Frame Resolution}}
\end{wrapfigure}
Dynamic Frame Resolution (DFR) trades rendering performance for visual
quality. The rendering for a channel is done at a different resolution
than the native channel resolution to keep the framerate constant. The
DFR equalizer adapts the zoom of a channel, based on the target and
current framerate. It is typically used for fill-rate bound
applications, such as volume rendering and raytracing.

\fig{fDFR}\footnote{Data set courtesy of Olaf Ronneberger, Computer
  Science Institute, University of Freiburg, Germany} shows DFR for
volume rendering. To achieve 10 frames per second, the model is rendered
at a lower resolution, and upscaled to the native resolution for
display. The rendering quality is slightly degraded, while the rendering
performance remains interactive. When the application is idle, it
renders a full-resolution view.

@@

Equalizer uses a C++ programming interface. The API is minimally
invasive, so Equalizer imposes only a minimal, natural execution
framework upon the application. It does not provide a scene graph, or
interferes in any other way with the application's rendering code. The
restructuring work enforced by Equalizer is the minimal refactoring
needed to parallelize the application for rendering.

All rendering threads in the configuration run asynchronously to the
application's main thread. Depending on the configuration's latency,
they can fall $n$ frames behind the last frame finished by the
application thread. A latency of one frame is usually not perceived by
the user, but can increase rendering performance substantially since
operations are pipelined better.

@@

In this section the source code of \textsf{eqPly} is explained in
detail, and relevant design decisions, caveats and other remarks are
discussed.

@@

The distributed data classes are helper classes based on
\textsf{eq::Object} or its parent class \textsf{eq::net::Object}. They
illustrate the typical usage of distributed objects for static as well
as dynamic, frame-specific data. Furthermore they are used for a basic
scene graph distribution of the model data.

@@

The second step is to redirect all Equalizer output to a file if this
was requested through the command line. This should happen before
initializing Equalizer, so that output during initialization is
redirected and threads launched by Equalizer will use the given log
stream. Any \textsf{std::ostream} can be used as log output:

{\footnotesize\begin{lstlisting}
    // 2. Redirect log output to file, if requested
    std::ofstream* logFile = 0;
    const std::string& logFilename = initData.getLogFilename();
    if( !logFilename.empty( ))
    {
        logFile = new ofstream( logFilename.c_str( ));
        eq::base::Log::setOutput( *logFile );
    }
\end{lstlisting}}

The third step is to initialize the Equalizer library. The
initialization function of Equalizer also parses the command line, which
is used to set certain default values based on Equalizer-specific
options\footnote{Equalizer-specific options always start with -\,-eq-},
e.g., the default server address. Furthermore, a \textsf{NodeFactory} is
provided. The \textsf{EQERROR} macro, and its counterparts
\textsf{EQWARN}, \textsf{EQINFO} and \textsf{EQVERB} allow selective
debugging outputs with various logging levels:

@@

The fourth step is to create an instance of the application and to
initialize it locally. The application is a subclass of
\textsf{eq::Client}, which in turn is an \textsf{eq::net::Node}. The
underlying Equalizer network layer, discussed in \sref{sNetwork}, is a
peer-to-peer network of \textsf{eq::net::Node}s. The client/server
concept is implement the higher-level \textsf{eq} client namespace.

@@

{\footnotesize\begin{lstlisting}
    // 4. initialization of local client node
    RefPtr< eqPly::Application > client = new eqPly::Application( initData );
    if( !client->initLocal( argc, argv ))
    {
        EQERROR << "Can't init client" << endl;
        eq::exit();
        return EXIT_FAILURE;
    }
\end{lstlisting}}%>>

@@

{\footnotesize\begin{lstlisting}
    // 5. run client
    const int ret = client->run();
\end{lstlisting}}

@@

{\footnotesize\begin{lstlisting}
    // 6. cleanup and exit
    client->exitLocal();
    client = 0;

    eq::exit();

    if( logFile )
        logFile->close();
    delete logFile;

@@

\begin{enumerate}
\item \textbf{Application}: The executable started by the user, the
  controlling entity of the rendering session.
\item \textbf{Auto-launched render client}: The typical render client,
  started by the server. The server starts the executable with special
  parameters, which cause \textsf{Client::initLocal} to never
  return. During exit, the server terminates the process. By default,
  the server starts the render client using \textsf{ssh}. The launch
  command can be used to configure another program to auto-launch render
  clients.
\item \textbf{Resident render client}: Manually pre-started render
  client, listening on a specified port for server commands. This mode
  is selected using the command-line option \textsf{--eq-client} and
  \textsf{--eq-listen \textless address\textgreater} to specify a
  well-defined listening address, and potentially \textsf{-r} to keep
  the client running across multiple runs\footnote{see
    \link{http://www.equalizergraphics.com/documents/design/residentNodes.html}}.
\end{enumerate}

@@

Another configuration can be selected using the command-line parameter
\textsf{--eq-config}. This configuration may only contain an
\textsf{appNode}, i.e., no distributed configuration is possibly since
the process-local server does not create a listening socket.

@@

When a new frame is started, the server computes all rendering tasks and
sends them to the appropriate render client nodes. The render client
nodes dispatch the tasks to the correct node or pipe thread, where they
are executed in order of arrival.

\textsf{Config::finishFrame} blocks on the completion of the frame
\textsf{current - latency}. The latency is specified in the
configuration file, and allows several outstanding frames. This allows
overlapping execution in the node processes and pipe threads and
minimizes idle times.

By default, \textsf{Config::finish\-Fra\-me} also synchronizes the
completion of all local rendering tasks for the current frame. This
facilitates porting of existing rendering codes, since the database does
not have to be multi-buffered. Applications such as \textsf{eqPly}, which
do not need this per-node frame synchronization, can disable it as
explained in \sref{sThreadModel}.

\begin{figure}[ht!]\center
  \includegraphics[width=\textwidth]{images/syncAsync}
  {\caption{\label{fSyncAsync}Synchronous and asynchronous execution}}
\end{figure}
\fig{fSyncAsync} shows the execution of the rendering tasks of a 2-node
2D compound without latency and with a latency of one frame. The
asynchronous execution pipelines certain rendering operations and hides
imbalances in the load distribution, resulting in an improved
framerate. For example, we have observed a speedup of 15\% on a
five-node rendering cluster when using a latency of one frame instead of
no
latency\footnote{\link{http://www.equalizergraphics.com/scalability.html}}. A
latency of one or two frames is normally not perceived by the user.

@@

If the application detects that it is currently idle, all pending
commands are gradually flushed, while still looking for a redraw
event. Then it waits and handles one event at a time, until a redraw is
needed:

@@

        while( !config->needsRedraw( )) // wait for an event requiring redraw
        {
            if( hasCommands( )) // execute non-critical pending commands
            {
                processCommand();
                config->handleEvents(); // non-blocking
            }
            else  // no pending commands, block on user event
            {
                const eq::ConfigEvent* event = config->nextEvent();
                if( !config->handleEvent( event ))
                    EQVERB << "Unhandled " << event << endl;
            }
        }
        config->handleEvents(); // process all pending events
    }
\end{lstlisting}}%>>

@@

Equalizer provides distributed objects which facilitate the
implementation of data distribution in a cluster
environment. Distributed objects are created by subclassing from
\textsf{eq::Object} or \textsf{eq::net::Object}. The application
programmer implements serialization and deserialization of the
distributed data.

Objects are dynamically assigned to a session, which assigns
session-unique identifiers to objects and manages the network
communication between different instances of the same distributed
object.

The master version of a distributed object is registered with a
\textsf{eq::net::Session}, which assigns a session-unique identifier to
the object. This identifier can be used by other nodes to map their
instance of the object, thus synchronizing the object's data with the
remotely registered master version.

Distributed objects can be static (immutable) or dynamic. Dynamic
objects are versioned.

@@

\begin{wrapfigure}{r}{.382\textwidth}
%\begin{figure}[h!t]
  \vspace{-2ex}
  \subfigure[]{\label{fObjectDist1}
    \includegraphics[width=.382\textwidth]{images/umlObjectSubclass}}\hfil
  \subfigure[]{\label{fObjectDist2}
    \includegraphics[width=.382\textwidth]{images/umlObjectProxy}}\hfil
  \subfigure[]{\label{fObjectDist3}
    \includegraphics[width=.382\textwidth]{images/umlObjectMI}}%
  {\caption{\label{fObjectDist}Object distribution using subclassing,
      proxies or multiple inheritance}}
  \vspace{-4ex}
\end{wrapfigure}
Distributed objects are addressed using session-unique identifiers,
because pointers to other objects cannot be distributed directly; they
have no meaning on remote nodes. The session used to register
distributed objects is normally the \textsf{eq::Config}, which is
subclassed from \textsf{eq::net::Session}.

The entry point for shared data on a render client is the identifier
passed by the application to \textsf{Config::init}. This identifier
typically contains the identifier of a static distributed object, and is
passed by Equalizer to all \textsf{configInit} task methods. Normally
this initial object is mapped by the render clients in
\textsf{Node::configInit}. It typically contains identifiers of other
shared data objects.

The distributed data objects referenced by the initial data object are
often versioned objects, to keep them in sync with the rendered
frames. Similar to the initial identifier passed to
\textsf{Config::init}, an object identifier or object version can be
passed to \textsf{Config::startFrame}. Equalizer will pass this
identifier to all \textsf{frameStart} task methods. In \textsf{eqPly},
the frame-specific data, e.g., the global camera data, is versioned. The
frame data identifier is passed in the initial data, and the frame data
version is passed with each new frame request.

There are multiple ways of implementing data distribution for an
existing C++ class hierarchy:
\begin{description}
  \item[Subclassing]: The classes to be distributed inherit from
    \textsf{eq::net::Object} and implement the serialization
      methods. This requires modification of the existing classes. This
      is used for \textsf{eqPly::InitData} and
      \textsf{eqPly::FrameData}. (\fig{fObjectDist1})
  \item[Proxies]: For each object to be distributed, a proxy object is
    created which manages data distribution for its associated
    object. This requires the application to track changes on the object
    separately from the object itself. The model data distribution of
    eqPly is using this pattern. (\fig{fObjectDist2})
  \item[Multiple Inheritance]: A new class inheriting from the class to
    be distributed and from \textsf{eq::net::Object} implements the data
    distribution. This requires the application to instantiate a
    different type of object instead of the existing object, and to
    create wrapper methods in the superclass calling the original method
    and setting the appropriate dirty bits. This pattern is not used in
    eqPly. (\fig{fObjectDist3})
\end{description}

@@

Equalizer determines the way changes are to be handled by calling
\textsf{Object::get\-Change\-Type} during the registration of the master
version of a distributed object. The change type determines the memory
footprint and the contract for the serialization methods. The following
change types are possible:

@@

The data streams implement efficient buffering and data transport
between nodes. They perform no type checking or transformation on the
data. It is the application's responsibility to exactly match the order
and types of variables during serialization and de-serialization.

Custom data type serializers can be implemented by providing the
appropriate serialization functions. No pointers should be directly
transmitted through the data streams. For pointers, the corresponding
object is typically a distributed object as well, and it's identifier
and potentially version is transmitted in place of its pointer.

For \textsf{InitData}, serialization in \textsf{getInstanceData}
and de-serialization in \textsf{applyInstanceData} is performed by
streaming all member variables to or from the provided data
streams:

{\footnotesize\begin{lstlisting}
void InitData::getInstanceData( eq::net::DataOStream& os )
{
    os << _frameDataID << _windowSystem << _renderMode << _useGLSL << _invFaces
       << _logo;
}

void InitData::applyInstanceData( eq::net::DataIStream& is )
{
    is >> _frameDataID >> _windowSystem >> _renderMode >> _useGLSL >> _invFaces
       >> _logo;

@@

\subsubsection{Assembly}
